%!TEX root = ../plos_template.tex
Different gene regulatory network architectures are commonly thought to provide access to different collections of gene expression patterns that ultimately result in different phenotypes. The manner in which stochasticity in the gene expression process impacts the relationship between network architecture and the expression patterns each is capable of producing is not clear. In order to investigate this relationship, we begin with the finest grained notion of genotype to phenotype map, gene expression, and adopt a formalism to reason generally about probability distributions over such maps~\cite{Lane1998,MacLane1992,Awodey2006,Abramsky2011}. In all cases, this must be relative to a notion of network architecture, defined here as any hypergraph over a given set of genes. We show that in a simple case where the highest-order gene regulatory interactions are allowed, the collection of accessible phenotypes is smaller than some of the alternative cases in which only lower order interactions are allowed. This conclusion is arrived at by directly comparing the geometries of the spaces of probability distributions on genotype-phenotype maps for the case of a genome containing a given number of simultaneously interacting genes to the equivalent for all of the other possible gene regulatory network topologies on the same number of genes. We find that the latter spaces of probability distributions are larger than the former one whenever the gene regulatory network topology contains a cyclic structure. A generalization of this result to an arbitrary number of genes implies that modularization of interactions, defined here as restriction from relatively higher to lower order interactions, within and among gene regulatory networks permits access to additional correlation patterns among the expression of multiple interacting genes. That these additional gene expression patterns may enable the fulfillment of functional requirements in certain environments while remaining intrinsically inaccessible when highest-order interactions are allowed implies that even when gene expression stochasticity is taken into account network architectures may be differentiated via natural selection in a background of sizeable neutral spaces. The identification of \emph{regulatory constraint}~\cite{Bar-Even2006,Johnson2010a} as a necessary enabler, as opposed to an inhibitor, of \emph{phenotypic diversity} may contribute to a more general explanation for the hierarchical modular architecture~\cite{Ravasz2002,Segre2005,Wagner2007,Erwin2009,Jothi2009,Bhardwaj2010,Colm} of the gene regulatory networks and, more generally, interactomes observed throughout the tree of life.
